\documentclass[
  seminar,        % seminar | bsc2019 | msc2019 | msc2019bio | mscwirtschaft2019
  english,        % ngerman | english
  titlepage=on,
  captions=tableheadings
]{ttlab-qualify}

\addbibresource{literature.bib}

\titlehead{
  Henri von Weissenborn\\
  7987764\\
  Informatik (B.Sc.)\\
  Henri von Weissenborn
}

\subject{Text Analytics}
\author{Henri von Weissenborn}
\title{Titel der Arbeit}
\subtitle{Optionaler Untertitel}
\date{Abgabedatum: \today}

\publishers{
  Goethe-Universität Frankfurt\\
  Institut für Informatik\\
  Text Technology Lab\\
  Betreuer: Name
}

\begin{document}

\maketitle
\tableofcontents
\clearpage

%\label{ch:introduction}

\section{Problemstellung}
With the rapid advancement in Large Language Models (LLMs) the human-computer interaction is fundamentelly transfromed. Models like ChatGPT, Llama and Claude are capable of producing text that is  grammatically fluent, stylistically coherent and basically indistinguishable from human writting. However, this surface-level fluency hides a fundamental flaw. LLMs frequently generate content that on first sight appear plausible due to the convincing nature but is either factually incorrect or unsupported by the provided source material. Literature refers to this phenomena as \textit{hallucination}. 
As AI and LLMs are more and more incorporated in everydays life raising concers come up about the trustworthiness of these new technologies. Esspecially the deployment in high stakes domains such as the medical sector, legal consultation or in educational support reliablity and safety are critical. The consequences of such hallucinations can range from minor inaccurancies to serious misinformation impacting human lifes. What makes this problem particularly difficult to address is that hallucinated content is not always obviously wrong. A model may confidently cite a non-existing study, subtly misattribute a quote or generate a translation unfaithful to the source. The plausibility 
of hallucinated content is precisely what makes it dangerous. Users 
who lack domain expertise or access to the source material have no 
reliable means of detecting such errors on their own. Motivated by these risks, recent work emphasizes the need for automated detection and evaluation methods, especially in scenarios where human verification is infeasible (e.g., cross-lingual use cases where users cannot reliably assess translation correctness). 

% Motivation

\subsection{Seed Paper and Thesis}

Building upon the challenges outlined above, this paper is centered 
around the seed paper \textit{"Can LLMs Detect Intrinsic Hallucination 
in Paraphrasing and Machine Translation"} 
(\cite{Gogoulou2025LLMHallucinationDetection}), which investigates the 
inherent ability of LLMs to detect intrinsic hallucination in two 
scenarios: paraphrasing and machine translation. The paper frames this 
detection task as an entailment-based reasoning problem and systematically 
compares LLMs against Natural Language Inference (NLI) models, examining 
how factors such as model size, instruction-tuning, prompt language, and 
prompt formulation influence detection performance.

The central thesis of this paper is to examine deeper \textit{how well LLMs 
are suited for detecting intrinsic hallucinations}. To 
address this question, the paper first establishes the theoretical 
groundwork by defining all necessary taxonomy Section~\ref{sec:Definitions and theoretical basis}. Section~\ref{sec:approaches} 
then reviews alternative detection approaches and discusses their 
relevance for future hallucination detection systems. Building on 
these findings, Section~\ref{sec:discussion} critically examines 
whether LLMs are inherently capable of reliably detecting intrinsic 
hallucinations, before the paper concludes by outlining which 
approaches offer the most promise for future development.

\section{Definitions and theoretical basis}

\subsection{Hallucination}

In the real world, hallucination is defined as \textit{the perception of an entity or event that is absent in reality} (\cite{MacphersonPlatchias2013Hallucination}, S.TOO). In the context of Language generation, hallucination refers to a models output that on first sight seem fluent and plausible yet is either factually incorrect or not supported by a given source or context. In general does the specific definition of hallucination vary between studies yet this a common characterization.\cite{Ji2023HallucinationSurvey} defines hallucination as \textit{generated content that is either nonsensical or unfaithful to the provided source content} which is widely adopted in the research community. It is therefore important to note that the concrete boundaries of what counts as hallucination are not universally defined. Different research communities and tasks apply the term with different weight and strictness. Hallucination substantially depends on the application context, a point which will be mentionend and discusses in greater detail in the following sections.

\subsection{Intrinsic vs. Extrinsic Hallucination}

In the context of LLMs, hallucination is commonly further divided into intrinsic and extrinsic hallucination.

\newline
\textbf{Intrinsic hallucination} generally describes cases where the output is deficient with respect to a particular input and or contradicts it. Crucially, can this deficiency be detected solely by comparing the input and output, without requiring external world knowledge (\cite{Ji2023HallucinationSurvey}, S.4). (\cite{Ji2023HallucinationSurvey}, p.4) illustrates this clearly: given the 
source sentence \textit{"The first vaccine for Ebola was approved by the 
FDA in 2019"}, a model generating the summary \textit{"The first Ebola 
vaccine was approved in 2021"} produces an intrinsic hallucination, as 
the date directly contradicts the source. Similarly, in a machine 
translation context, translating the German sentence \textit{"Michael 
ging am Donnerstag in den Buchladen"} as \textit{"Jerry didn't go to the 
bookstore"} leads to an intrinsic hallucination, since both the 
subject and the negation contradict the original input. The inconsitency is fully self-contained within the input-output pair and can in principle be detected through semantic consistency reasoning alone. 



\textbf{Extrinsic hallucination} on the other hand, describes cases, that involve information that is not present in the input and cannot be verified solely based on the source alone. In order to detect extrinsic hallucination it typically requires an external knowledge source. Crucially it does not matter whether the part which is not entailed happens to be factually correct. Extrinsic hallucination can therefore still be accurate compared to world knowledge. 

An example from \cite{Ji2023HallucinationSurvey} shows the concrete distinction between intrinsic and extrinsic hallucination. Looking at the same source text about the Ebola vaccine, 
a model generating the summary \textit{"China has already started 
clinical trials of the COVID-19 vaccine"} will produce an extrinsic 
hallucination because neither the source nor any internal reasoning confirms or denies this claim. Likewise, for the machine translation scenario, 
translating \textit{"Michael ging am Donnerstag in den Buchladen"} as 
\textit{"Michael happily went to the bookstore with his friend"} 
constitutes an extrinsic hallucination. Adding the extra details 
\textit{"happily"} and \textit{"with his friend"} introduce new information which the source cannot verify, even though the translation 
is not directly contradictory. 
\newline

\subsection{Faithfulness, Factuality and Entailment}
Terms also often used in the context of hallucination are the factuality, faithfulness and entailment. Faithfulness describes how well a model's generated output is consistent by the given input or user provided context. Therefore faithfulness is closely related to intrinsic hallucination. (\cite{Ji2023HallucinationSurvey}) describes it as \textit{staying consistent and truthful to the provided source - antonym to "hallucination"}. Thereby if we try to maximize faithfulness of an output it should contain no intrinsic hallucination. As intrinsic hallucination does not require any outside knowledge, faithfulness-oriented evaluation methods or other interference models are particular well suited for this detection task. Especially in conditional generation tasks such as summarization, paraphrasing or machine translation. 
\newline
On the contrary \textit{Factuality} refers to the quality of being actual or based on fact. However does the interpretation of the "fact" vary. Papers like (\cite{Dong2020MultiFactCorrection}) treat the source itself as the factual reference, effectively given the term "factuality" and "faithfulness" the same meaning. For both the seed paper and this thesis, we adopt the definition proposed by \cite{Maynez2020FaithfulnessFactuality}, which clearly distinguishes source knowledge from world knowledge.
\newline


\subsection{detection instead of mitigation}
An increasing part of recent research focuses on hallucination mitigation, meaning that at the time of generation the output is being checked for hallucination. However, in many practical cases this is not enough, due to inherent problems such as biases, knowledge gaps, and architecture which are fundamentally embedded in the model (\cite{Ji2023HallucinationSurvey}, S. 42:19). Hallucination detection is therefore not merely a complement to mitigation but an independent research problem in its own right. Mitigation on one hand can leverage its internal states or generation process to intervene during generation.
Detection on the other hand must solely reason post-hoc from the final input-output pair without access to the model that generated the text.
Especially for the detection of hallucination, this means the detector must perform stable semantic consistency reasoning under no addional constraints. The following sections will show why detection poss no trivial problem.

% Term faithfulness varies from task to task Survey ZI p.6
\subsection{Intrinsic Hallucination as a Non-Trivial problem}
While intrinsic hallucination may appear to be a straightforward consistency check between source input and model output, both empirical evidence and theoretical analysis reveal that it constitutes a fundamentally non-trivial problem.
On the empirical side, even for humans the JUDGE-BENCH (\cite{tan2024judgebench} study shows varying agreement not only between humans and models, but also among human annotator themself. The reports in Table 3 (pp.11-12), show that inter-annotator agreement (e.g. Krippendorff's alpha between humans vary strongly for certain tasks, indicating that even humans frequently disagree on properties like coherence, consistency and faithfulness. These results suggest that intrinsic hallucination may not be a binary or objectively trivial problem but often requires interpretation. Especially in cases of faithfulness and entailment it is not trivial to determine whether a text should be considered to contain hallucination or not. 

 Furthermore, does \cite{Ji2023HallucinationSurvey} note that the tolerance for what constitutes hallucination varies across tasks and domains: what is considered a faithful output in an open-domain dialogue system may be classified as intrinsic hallucination in a summarization or machine translation context. This definitional instability further adds to the difficulty of detection, as evaluation methods must be calibrated not only to a model's output but also to the circumstances of the specific task. Taken together, these considerations establish that intrinsic hallucination detection is neither a solved problem nor a simple classification task, but a challenge that operates simultaneously at the level of linguistic consistency, human judgment, and mathematical computability. These limitations motivate a more detailed investigation and will be discussed in the subsequent sections of this paper. The following section focuses on intrinsic hallucination detection and examines how this problem is addressed in the seed paper Can LLMs Detect Intrinsic Hallucinations in Paraphrasing and Machine Translation?

\section{Seed Paper at the Core: Detecting Intrinsic Hallucinations}
This section presents the seed paper "Can LLMs Detect Intrinsic Hallucination in Paraphrasing and Machine Translation?" as the core of this seminar paper. The paper focuses on intrinsic hallucination and explores whether LLMs can reliably detect such errors.
\newline
\subsection{Methods of paper}
The seed paper formulates intrinsic hallucination as an entailment problem. Therefore is an output faithful if it is fully entailed by the source input and considered to contain hallucination if it contains additional or contradictory information not logically supported by the source. This formulation allows the detection task to be treated as a reasoning task rather than a knowledge verification problem. To conduct this approach, the authors construct a contrastive evaluation setup. Each instance consists of a source sentence and two candidate hypotheses, where always one is faithful and the other contains intrinsic hallucination. The model is then asked to determine which of the hypothesis is the hallucinated one. This setup ensures a controlled environment and allows for direct comparison across models. The methodology is based on the HalluciGen dataset TODO CITE, which this paper is based upon and specially designed for intrinsic hallucination detection. The dataset convers two szenarios: paraphrasing (PG) and machine translation (MT). For the paraphrasing task, both source and hypotheses are in the same language while the translation task involves cross-lingual input and output. Additionally did the paper test different approaches to prompt the LLM. The study first analyzes the effect of prompt language, testing whether alignment between the prompt language and the language of the source and hypotheses influences performance. In addition, it systematically compares different prompting styles to evaluate how prompt formulation shapes detection accuracy. Prompts (1-3) contain the term "hallucination" while prompts (3-6) only contain a explicit definition of the concept. Prompt (4-6) on the other hand contain neither and uses formulations that describe hallucination. For example, Prompt 4 is formulated as:  \textit{"Given a source sentence (src) and two <scenario> hypotheses (hyp1 and hyp2), detect which one of the two logically contradicts the src."} (\cite{Gogoulou2025LLMHallucinationDetection}, S. 5). Lastly to optimize outputs and ensure proper formatting in addition to the prompts every model receives the same instructions. Models are instructed to only answer with either "hyp1" or "hyp2". TDO CITE modelle die getestet wurden
\subsection{Results of the Paper}
The experiments of \cite{Gogoulou2025LLMHallucinationDetection} yield three findings central to the thesis of this paper. First, LLM performance varies considerably across models, but the strongest models — META-LLAMA-3-70B-INSTRUCT and MIXTRAL-8X7B-INSTRUCT — achieve competitive F1 scores across both tasks and languages, demonstrating that large, instruction-tuned LLMs have a measurable inherent capacity for intrinsic hallucination detection. Second, and most critically for this thesis, the NLI baseline BGE-M3-ZEROSHOT-V2.0 achieves F1 scores of 0.90 and 0.92 on English and Swedish paraphrase respectively, outperforming all LLMs in the paraphrase scenario. This result challenges the assumption that LLMs are the superior or necessary choice for this task. Third, none of the four investigated factors — model size, instruction tuning, prompt language, and prompt formulation — can serve as a reliable predictor of performance, as their effects are inconsistent across model families. Taken together, these findings suggest that LLMs are \textit{conditionally} suited for intrinsic hallucination detection: capable under the right conditions, but neither universally reliable nor inherently superior to simpler entailment-based approaches.


\section{Alternative Approaches to Instric Hallucination detection}
Although the seed paper introduces a specific investigation into the detection of intrinsic hallucination using LLMs and NLI models, it represents only a part of a broader area of research. To better classify these results this section will review alternative approaches and compare them with the entailment-based framework adopted in this seed paper.

\subsection{Entailment-Based Approaches}
The core assumption of intrinsic hallucination is that faithful outputs must be entailed by their source input. Other studies also showed promising results of the capabilities of NLI models for such tasks. Henceforth, has Natural Language Inference (NLI) emerged as a particular fitting system for detecting such inconsistencies. Early studies, such as \cite{Maynez2020FaithfulnessFactuality}, showed that pretrained NLI models can reliably identify inconsistencies in abstractive summarization. 

This entailment-based framing is a central point in the seed paper \cite{Gogoulou2025LLMHallucinationDetection}, which explicitly describes intrinsic hallucination detection as a controlled NLI problem. The findings therefore reinforce the view that intrinsic hallucination detection is fundamentally a semantic consistency problem. Therefore are models, that are capable of entailment reasoning a promising solution. The good results of NLIs come with no suprise as the must “handle phenomena like lexical entailment, quan-
tification, coreference, tense, belief, modality, and lexical and syntactic ambiguity” (Williams et al., 2018) to successfully predict entailment, contradiction, and neutral relations between sentence pairs. However, \cite{mishra2021react} highlight limitatiations of NLI-based hallucination detection. Standard NLI models are typically trained on sentence-pair datasets. Although Mishra et al. (2021) conduct their empirical analysis in the context of abstractive summarization, the identified limitation of NLI-based evaluation—namely the granularity mismatch between sentence-level entailment training and document-level generation tasks—extends beyond summarization and affects other conditional generation settings as well.
% SCHREIBE ÜBER NLI UND ENTAILMENT

Beyond the entailment-based formulation used in the seed paper using NLIs, other research proposes different approaches for hallucination detection. The survey (\cite{Alansari2025LLMHallucinationSurvey} formulates different categories for detection methods. While the proposed methods were not originally made for intrinsic hallucination detection, several suit its definiton and are relevant when hallucination is framed as a violation of faithfulness. 

\subsection{Self-Consistency and Uncertainty-based Methods}
Another approach is self-consistency-based-detection. For example, (\cite{Manakul2023SelfCheckGPT}) proposes SelfCheckGPT, a framework combining LLM prompting, token-level probability signals and NLI-based entailment checks to detect hallucination on a sentence-level. The main idea of this approach is to generate multiple reponses for the same input evaluate the internal agreement. If the outputs of the model contradict another or vary at certain parts, this instability may imply hallucination. Although the paper does not explicitly differentiate between intrinsic and extrinsic hallucination, its reliance on entailment checks between source and output and the avoidance of external knowledge comparisons makes it particularly applicable for intrinsic inconsistencies. This work is closley related to uncertainty-based and LLM-based self-evaluation approaches. These methods assume that hallucinated content occurs from higher predictive uncertainty. These uncertainties can be then measured through entropy, token-level log probabilites or variance across multiple generations. Nevertheless does this approach has its flaws as it deduces that high confidence collelates with faithfulness. This may lead to cases where small logical errors like negation shifts, incorrect entities or numerical errors stay undected due to the high confidence of the model. 

% GPT
Semantic-based uncertainty methods aim to detect hallucinations by modeling uncertainty at the level of meaning rather than relying solely on token probabilities. Farquhar et al. [89] introduce semantic entropy as a measure for hallucination detection. Their approach generates multiple responses to the same prompt and clusters these outputs based on semantic similarity. Entropy is then calculated across the resulting semantic clusters. High semantic entropy indicates that the model’s responses differ substantially in meaning, suggesting a higher likelihood of hallucination. Conversely, low semantic entropy reflects consistent semantic content across generations and therefore greater reliability.

\subsection{Embedding-based Detection}
Embedding-based hallucination detectoin relies on measuring the semantic similiarity between output and the models input or source reference. The underlying assumption is that a faithful output should remain semantically close to its source. If the model generated output differs in the embedding space it could mean a semantic shift happend and therefore it contains hallucination. These approaches are typically further divided into seperate categories. \textbf{Similarity-based methods} compares sentences or embeddings of the source and output. (cite dale et al) for example used cross-lingual embedding models such as LaBSE to detect hallucination in machine translation. \textbf{Gradient-based Methods} additionally incorparates the gradient of information measuring how sensentive outputs depend on the input. For example, (\cite{Hu2024Embedding}, S.1954) utilize embedding shifts with uncertainty signals to further imporove detection performance. Overall, embedding-based methods are useful for detecting semantic discrepancies which can be useful for Paraphrasing or Machine Translation tasks. 

\subsection{Prompt-based Approaches}
Prompt-based hallucination detection utilizes the exceptional instruction-following capabilities of LLMs by specifically prompting them to evaluate the faithfulness of generated texts. Instead of solely using uncertainty signales or similarty scores, recent studies, like the seedpaper (\cite{Gogoulou2025LLMHallucinationDetection}, used LLMs as the evaluation instance. One key advantage of this approach is the flexiblity. Using LLMs requires no addional training and can generalize across several tasks and languages. \cite{Gogoulou2025LLMHallucinationDetection} showed that in a contrastive setting LLMs have the inherent ability to determine a hallucinated outcome. While this scenario may archieve decent results in the experiment using LLMs-as-the-Judge also brings its flaws. As shown does the effectivness heavily depend on the model, prompt formulation and the degree of instruction tuning. Additionally, the optimized experimental setup simplifies the generall task by given the LLM a choice between two possible hypothesis. This contrastive framing reduces
TODO GUCKEN OB AUSFÜHREN BZGL. LIMITATIONEN
%- Binär oder Likert Scale als evaluationsmetric
%- chain of thought prompting, allowing models to produce evalutaion reuslts along with explanations
%- Shift promblem point out problematic areas instead of classifying whole file

% Human annotators and LLMs might have the same biases JUDGE-BENCH / Limitations


% Finally, as shown in Figure 4, all models achieve better alignment with human judgments when
evaluating human language than when assessing
machine-generated text, both for categorical and
graded annotations. This result aligns with the find-
ings by Xu et al. (2024), suggesting that LLMs
display a bias towards their own generation. More
broadly, this trend calls for caution when using
LLMs to automatically evaluate the output of NLP
systems.
%%%%
\subsection{Important features detection hallucination / conclusion}
Taken together, these alternative approaches demonstrate that intrinsic hallucination detection can be addressed through lexical alignment, semantic entailment modeling, probabilistic confidence estimation, or meta-evaluation via prompting. However, each method captures different dimensions of inconsistency and exhibits distinct limitations. This reinforces the insight that intrinsic hallucination detection is fundamentally a problem of semantic consistency rather than merely confidence estimation, and it highlights why entailment-based frameworks, such as the one employed in the seed paper, provide a particularly principled foundation for analyzing model behavior.


\section{Discussion: Are LLMs Suited for Detection?}

\subsection{LLM-as-Judge Perspective}
% HUMANS ARE BAD ANNOTNATOR / NLP STUDY

Building upon the idea of using LLMs as a detection method for intrinsic hallucination, a 
more fundamental question arises: \textit{are LLMs inherently suited to detect hallucination?} 
To address this, it is useful to situate the detection task within the broader paradigm of 
LLM-as-a-Judge.

At its core, LLM-as-a-Judge refers to the use of large language models as automated evaluators 
to assess texts for criteria like quality, correctness, or in our case faithfulness. Before the 
era of LLMs, evaluation of natural language processing was driven by human evaluations which 
were costly, slow and difficult to scale, while automatic metrics such as BLEU or ROUGE 
provided consistency yet failed to properly capture deeper semantics. The increasing ability 
of LLMs to mimic human reasoning now enables automated evaluation across a wide range of tasks 
and domains without the bottleneck of human annotation (\cite{Gu2025LLMJudgeSurvey}, S.2). 
Another key advantage lies in their broad generalization capabilities: instruction-tuned LLMs 
have been trained on many diverse tasks and domains, allowing them to apply learned criteria 
across new tasks without requiring task-specific fine-tuning. \cite{Gogoulou2025LLMHallucinationDetection} 
showed that a well-designed prompt can be sufficient to guide a model towards producing 
structured judgments in a contrastive detection task. This inherent ability to mimic human 
judgment in addition to the flexible adaptation to new tasks shows that LLMs are generally 
suited for annotation tasks. 

The question that follows is whether this general suitability extends specifically to 
faithfulness evaluation. As dedicated research on LLMs judging faithfulness specifically is still limited, we draw on 
findings from the broader LLM-as-a-Judge literature. Findings about subjective quality 
assessment, for example \textit{consistency} or \textit{coherence} in 
(\cite{Bavaresco2025JudgeBench}), come close to faithfulness but cannot be transferred 
directly. However, structural weaknesses of LLMs as evaluation instances such as 
prompt sensitivity, sycophantic behavior, and the recursive problem of self-evaluation — are 
rooted in model architecture rather than task definition. They therefore apply to the detection role as well, regardless of whether the task is subjective or formally defined.

Despite the general advantages of LLMs as evaluators, adopting them as a replacement for 
human evaluators introduces a wide range of new challenges which cannot be overlooked. One 
major concern is reliability, as it directly impacts the robustness, stability and credibility 
of the generated judgment (\cite{Gu2025LLMJudgeSurvey}, S.6). The following subsections 
examine these structural weaknesses and their specific implications for intrinsic hallucination 
detection.


\subsection{Systematic biases}

Systematic biases rooted in the training process. A further weakness stems from how LLMs are trained. (Alansari and Luqman, 2025, p. 8) identify two particularly relevant failure modes at the fine-tuning stage: capability misalignment, where models are trained to give confident answers even when their knowledge is insufficient, and sycophantic behavior, where models generate outputs that appear plausible and evaluator-pleasing rather than factually grounded. Both tendencies directly undermine the reliability of LLMs as hallucination detectors: a model trained to appear confident and coherent will systematically underreport uncertainty and may favor plausible-sounding outputs over faithful ones. This concern is reinforced by (Bavaresco et al., 2024, S. 241), who find that LLM judges correlate more strongly with non-expert human annotations than with expert judgments, suggesting that LLMs may be approximating surface believability rather than performing genuine faithfulness evaluation. 

Prompt sensitivity as an instability indicator. The seedpaper (Gogoulou et al., 2025) provides direct empirical evidence of a further weakness: the sensitivity of LLM performance to prompt formulation. MIXTRAL-8X22B-INSTRUCT shows a standard deviation of up to 0.23 across prompt variations in the Swedish paraphrase scenario — a range so large it renders the model unreliable for this task. Notably, including the term "hallucination" in the prompt negatively impacts performance for several models, suggesting that the models' internal representations of the concept do not align consistently with its operational definition in the detection task. This prompt sensitivity is not a minor engineering concern: it reveals that LLM-based detection is not a stable, principled evaluation procedure but one whose results depend heavily on framing choices external to the model itself. (Alansari and Luqman, 2025, p. 1) arrive at a structurally similar conclusion from a broader perspective, noting that no single detection approach performs well under all circumstances and that the inconsistency in prompting techniques is itself a barrier to developing universal detection frameworks.




\subsection{Theoretical Limitations}
% LLMS können selber halluzinieren

A recurring assumption in prompt-based hallucination detection is that LLMs, due to their broad semantic capabilites, are in reality well-suited to evaluate the faithfulness of a generated text. However, at a closer look at LLMs reveal multiple systematic flaws that could fundamentally impact reliable detection.

The recursiv problem of self-evaluation. The most fundamental limiation with using LLMs as the hallucination detector is them being prone to same biases as the evaluate text itself. LLMs themself are generative probabilistic models and therefore susceptible to producing hallucinated content. (\cite{Banerjee2024LLMsHallucinate}, 2024, S. XX) argues, that hallucination is not an occasinal phenomena but an error which stems from the mathematical and logical structure of large language models in generell. Since LLMs work as an autoregressiv system, predicting the next token based on learned probability distributions and previous context, rather than learned truths, they inherently do not possess the ability to guarantee correctness. Henceforth if such a system is tasked to detect hallucination in an output, which is generated by itself or another model, a recursive problem emerges. 
(\cite{Banerjee2024LLMsHallucinate}, 2024) goes further and demonstrates that hallucination is a structural problem, which exists at every stage of the LLM pipeline. Incomplete training data, 
These proofs underline that even architectural improvements, larger datasets or post-hoc detection can only reduce but never fully eliminate hallucination. Even though Barnajee et al. primarly describes factual inconsistencies and therefore extrinsic hallucination the arguments still applies for intrinsic hallucionations, albeit in a slightly different form. As intrinsic hallucination evaluatues the faithfulness of an input in comparission to an output detecting such inconsistencies still requires reliable intent interpretation, accurate retrival of relevant source segments and first and foremost stable semantic comparisons between both texts. All of these properties are subject to uncertainty and if they cannot be entirely guaranteed due to the probabilistic nature of LLMs, evaluations can never be fully faithful. Even if the task is limited to only internal consistencies, the evaluation process itself remains grounded in probabilistic modelling and is therefore as shown in (\cite{Banerjee2024LLMsHallucinate}, 2024, p.20) mathematically unable to ensure perfect reliability.



% Gute zusammenfassung
\subsection{Conlusion LLMs for detection}
Taken together, the findings of the survey suggest a nuanced conclusion. LLMs are potentially suitable for hallucination detection because they combine semantic reasoning with scalability, outperforming shallow automated metrics . However, their suitability is conditional rather than absolute. Reliability depends on validated agreement with human experts, systematic bias mitigation, adversarial robustness testing, careful prompt engineering, and possibly ensemble integration. Moreover, because LLM judges can themselves hallucinate, they cannot guarantee objective truth verification.

Therefore, based  the survey’s framework, LLMs should not be viewed as autonomous truth detectors but as probabilistic evaluators whose effectiveness in hallucination detection depends on rigorous reliability engineering. They are suitable under controlled, validated conditions, but unsuitable as standalone, uncalibrated arbiters of factual correctness.

% lack of domain-specific model trianing data common phenomenon S.12 Survey 2026
%
\section{From Detection to Diagnosis}

\subsection{SOTA detection Methods}
Across the last years of research, no single system dominates intrinsic hallucination detection across all domains, tasks and languages. As reviewed in REFSECTION each method use different signals and therefore capture different dimensions of semantic inconsistency. However instead of using a single approach the best performing detection systems in the SemEval-2024 SHROOM shared task (\cite{mickus2024shroom}) showed that combining mulitple signals will yield the best results. This idea was also used by the winning system of this shared task MetaCheckGPT (\cite{mehta2024metacheckgpt}). Concretely, MetaCheckGPT proceeds in two stages. In the first stage, each candidate sentence is compared against a set of stochastically generated responses following the self-consistency principle of SelfCheckGPT (\cite{Manakul2023SelfCheckGPT}): multiple completions are sampled from models such as ChatGPT and Vectara, and each is asked whether it supports the original sentence. This creates a set of independent uncertainty signals for each base model used which reflects how consistently the sentence is supported across samples. In the second stage, a meta-regressor, aggregates all these signales in a single hallucination score. The key idea behind this approach is that a diverse panel of models capture a greater range of hallucination-indicative features than a single model or knowledge base. On the SHROOM test set, this approach archieved an accuracy of 0.847 on the model agnostic track (\cite{mickus2024shroom}, p. 1985).
The success of MetaCheckGPT is not an isolated finding but reflects the broader pattern of the systems used in this shared task. Hybrid systems combining complemantary detection signals consistently outperformaed single-method approaches. 
LLMs thus occupy an ambiguous position in this landscape. As prompt-based judges they offer unmatched flexibility and strong out-of-the-box performance, yet as shown in the seed paper 
(\cite{Gogoulou2025LLMHallucinationDetection}), their effectiveness depends heavily on model 
size, instruction tuning, and prompt formulation. The practical state of the art is therefore not a single model but a mixed architecture. 

\subsection{Limitations of the Classification Paradigm}

Despite these advances, all systems described above share a common architectural commitment: they frame hallucination detection as a classification problem, assigning a discrete label or a continuous risk score to a model output. This framing, while enabling benchmarking and automated filtering, is subject to a set of principled limitations. 

The most fundamental is the \textbf{absence of causal explanation}. A binary label flags that something is wrong but cannot communicate \textit{what} is wrong, \textit{where} the inconsistency occurs, or \textit{why} the model produced it. In practice, this severely limits the utility of detection for downstream purposes. A developer correcting a generative model can not act on a single label alone while a human reviewer examining a flagged output still must perform the diagnostic work manually.

A second limitation is \textbf{the lack of feedback for model improvement}. Detection systems are typically applied post-hoc to finished outputs. Even when a detector achieves high accuracy, it provides no gradient signal and no structured information that could be fed back into the generation process to reduce future hallucinations. This also represents a conceptual ceiling that detection can filter outputs but cannot cure the underlying failure mode (\cite{liu2025detection}).

Third, as established in Section~\ref{sec:Definitions}, \textbf{annotation ambiguity} means that the binary label itself is epistemically unstable. The JUDGE-BENCH study shows that inter-annotator agreement varies strongly across tasks (\cite{tan2024judgebench}), and the SHROOM task organisers explicitly warn that even top-performing systems can behave close to random on harder items (\cite{mickus2024shroom}, p.1985). Compressing a complex semantic consistency judgment into a single label therefore discards these uncertainties.

Finally, \textbf{prompt sensitivity} — extensively documented in the seed paper (\cite{Gogoulou2025LLMHallucinationDetection}) and the broader LLM-as-a-Judge literature (\cite{Gu2025LLMJudgeSurvey}) — means that the classification decision itself can be an artefact of framing rather than a stable property of the output being evaluated. A model that achieves F1 of 0.91 under one prompt formulation and collapses to near-random performance under another has not truly solved the underlying detection problem.

\subsection{The Diagnosis Framework}

Building upon these limitations the paper (\cite{liu2025detection}, 2025) proposes to expand the hallucination detection task to a Diagnosis task. The hallucinations diagnosis modell should therefore have the ability to examine the task across four dimensions. 
\begin{enumerate}
    \item \textbf{Detection} --- accurately identifying whether the generated content aligns with the source context.
    \item \textbf{Localization} --- pinpointing the specific segments within the output that are hallucinatory.
    \item \textbf{Explainability} --- explaining why the identified content is inconsistent with the source.
    \item \textbf{Mitigation} --- proposing corrections or generating a revised version of the output that restores faithfulness to the source.
\end{enumerate}

To operationalize these four diagnostic dimensions, the authors propose 
a two-component methodology. First, the \textbf{Hallucination Diagnosis 
Generator (HDG)}, an automated data synthesis pipeline, is introduced to 
construct a large-scale, richly annotated training dataset from raw corpora. 
Second, building upon this synthesized data, the \textbf{Hallucination 
Diagnosis Model (HDM-4B-RL)}, a 4-billion-parameter model, is trained 
using Group Relative Policy Optimization (GRPO) to guide the model toward 
the desired diagnostic behavior across all four dimensions.

The experimental results demonstrate the effectiveness of this approach. 
On the hallucination detection task, HDM-4B-RL outperforms all previous 
specialized detection models, surpassing the prior state-of-the-art model 
Lynx-8B by 2.4\% in average F1 score despite being half the size. 
Furthermore, the model achieves an average F1 score of 79.65, exceeding 
powerful closed-source models such as GPT-4.1 (75.37 F1) while operating 
at roughly half the inference time of the 32B-scale models.

In the full hallucination diagnosis task, covering detection, localization, 
and mitigation, HDM-4B-RL demonstrates highly competitive performance 
against much larger general-purpose models. Operating as a lightweight 
end-to-end single-prompt model, it closely approaches the diagnostic 
quality of the 32B-scale Qwen3 model, while requiring only $1\times$ 
the runtime compared to the $4.9\times$ demanded by the pipeline-based 
32B approach. These results validate that a compact, specialized model, 
when trained on high-quality diagnostic data, can serve as an efficient 
and reliable alternative to large general-purpose models in hallucination 
analysis tasks.

(\cite{liu2025detection}, p.1) draws the four-dimensional framework analogous to a clinical diagnosis: a physician must not only determine that a pathology is present, but also locate it, characterise its cause, and propose a treatment, a diagnostic hallucination model must not merely flag an error but reason about it.


\subsection{Where LLMs Fit in the Diagnosis Framework}

The shift from detection to diagnosis re-positions LLMs in an important way. In the detection paradigm, LLMs are used primarily as black-box classifiers, and their suitability is constrained by the structural weaknesses discussed in Section~\ref{sec:discussion}: prompt sensitivity, sycophantic tendencies, and the recursive problem of self-evaluation. In the diagnosis paradigm, however, LLMs are not asked to produce a single label but to engage in structured, multi-step reasoning — a task that aligns far better with their demonstrated strengths in chain-of-thought inference and instruction-following.

This observation suggests that the most productive architecture for future hallucination analysis systems may be a \textbf{hybrid}: an NLI-based or lightweight classifier performing fast, stable first-pass detection, followed by an instruction-tuned LLM handling the interpretive steps of localisation and explanation. The classifier provides the reliability and consistency that LLMs lack under distributional pressure; the LLM provides the semantic depth and flexibility that discriminative models cannot supply. Such a cascade would combine the principled entailment reasoning that the seed paper identifies as the core of intrinsic hallucination detection with the diagnostic richness that classification alone cannot provide.

Detection, in this view, is not the endpoint of hallucination research but its starting point. The binary label establishes that a problem exists; diagnosis begins the work of understanding it.


\section{Conclusion}
% Intrin hallu als konsistenzproblem
% Entailment-basierte Ansätzt prinzipel geeignet
% LLms geeignet under bedingungen aber nicht absolut zu verlässig

% flexiblität begriff faithfulness
Due to the broadness of tasks with which LLMs are often
used for, there are several domains which have utilized unique
strategies for assessing faithfulness, with these methodologies
having been specifically chosen and optimised for the domain
at hand. Faithfulness is a crucial evaluation criterion for all
domains within the scope of this review, as users rely on the
model to provide information that is factually correct and true
to the original source otherwise problems will arise with
varying levels of severity depending on the criticality of the task
at hand.
It can be challenging to define faithfulness universally due
to the intricacies required across domains. For example,
summarization requires the output to be aligned with the source
material, which can be validated without the requirement of a
ground truth. Whereas an open question-answering (QA) task
requires a ground truth to be provided for the faithfulness to be
measured.

\section{Limitations}
A central challenge in research on intrinsic hallucinations in large language models (LLMs) lies in the limitations of currently available datasets and evaluation methods \cite{Gu2025LLMJudgeSurvey}, p.49. Many studies rely on relatively small or task-specific datasets, such as NLI-style benchmarks or datasets designed for particular tasks like summarization or machine translation \cite{Gogoulou2025LLMHallucinationDetection}. These datasets are often manually created and carefully curated for a specific evaluation scenario, which makes them well suited for controlled experiments but limits their ability to generalize to other tasks or domains. As a result, conclusions drawn from such datasets are frequently highly task-dependent and may not reflect the broader phenomenon of intrinsic hallucinations in LLM outputs.
Another limitation is the lack of a consistent and widely accepted taxonomy of hallucination types. Different studies adopt slightly different definitions and categorizations of hallucinations, which makes it difficult to compare results across benchmarks or to develop standardized evaluation frameworks. Furthermore, defining hallucination itself remains challenging, as the classification of a statement often depends on how the underlying context or source document is interpreted.
% Paper sind teils schon 2 Jahre alt und eventuell auch nicht mehr aktuell mit dem rapiden entwicklungen von LLMs
Finally, there is still no universally accepted metric for reliably measuring hallucinations. Many commonly used evaluation metrics show inconsistent alignment with human judgments across datasets, models, and generation settings. This inconsistency makes it difficult to establish robust benchmarks and complicates the comparison of different hallucination detection approaches.
It therefore needs more extensive research particular in intrinsic hallucinations


% ====================
% Optional
% ====================

%\chapter*{Abkürzungsverzeichnis}
%\addcontentsline{toc}{chapter}{Abkürzungsverzeichnis}

% ====================
% Anhang
% ====================

\appendix

\printbibliography

\end{document}
